Eco-evolutionary dynamics of finite populations from first principles

### 2023-02-27 MS_thesis.tex
Move examples to appendix
Update abstract

%The abstract goes here. IISER guidelines say it should be <= 250 words.
\thesisabstract{
Some central questions of population biology relate to the number of distinct variants of an entity that can emerge and coexist in a closed population. Examples include the study of standing genetic variation (alleles), polymorphisms (genotypes/phenotypes), and sympatric speciation (species). Historically, these questions have often been studied through phenomenological models formulated for infinitely large populations. In this thesis, I use ideas from statistical physics to analytically describe evolving populations from biological first principles. Starting from a density-dependent ‘birth-death process’ describing an arbitrary closed population of individuals with discrete traits, I derive a set of stochastic differential equations for how trait frequencies change over time. Well-known relations such as the replicator-mutator equation and Fisher’s fundamental theorem are recovered in the infinite population limit. I then extend this theory to one-dimensional quantitative traits through a ‘stochastic field theory’ that yields frameworks such as Kimura's continuum-of-alleles and adaptive dynamics as special cases. I make no assumptions other than density-dependence and the impossibility of infinite growth, and thus, my derivation is almost universally valid. The resultant equations reveal a new directional evolutionary force, ‘noise-induced selection’, that is particular to finite populations and has been overlooked in standard mathematical formulations of evolution. Noise-induced selection can reverse the direction of evolution predicted by infinite-population models, with implications for simulation studies and real world populations. My work thus derives a set of fundamental equations showcasing the common stochastic underpinnings of several formal structures in population biology and predicts a new evolutionary force unique to finite populations.
}

@@

\part{Theory \& Applications}
\chapter{Population dynamics from stochastic first principles}\label{chap_BD}
\input{ind_files/03_BD_processes}
\chapter{Stochastic field equations for the evolution of quantitative traits}\label{chap_infD_processes}
\input{ind_files/04_field_eqns}
\chapter{Examples and Applications}\label{chap_examples}
\input{ind_files/05_examples}

\part{Discussion \& Outlook}
\chapter{A unified view of population dynamics}\label{chap_unification}
\input{ind_files/06_unification_discussion}
%\chapter{Towards a stochastic evolutionary theory}\label{chap_stochasticity}
%\input{ind_files/07_stochastic_discussion}
% \chapter{}
% \input{ind_files/chapter_08}
% \chapter{} <-- uncomment and add as required
% \input{ind_files/chapter_09}

@@

%make sure the title is 'Appendix' and not 'Chapter'
\renewcommand{\chaptername}{Appendix} 
\chapter{Deriving the Fokker-Planck equations for It\^{o} SDEs}\label{App_SDE_FPE}
\input{ind_files/99_Appendix_A}
\chapter{Deriving stochastic trait frequency dynamics using It\^{o}'s formula}\label{App_density_to_freq}
\input{ind_files/99_Appendix_B}
\chapter{A Price-like equation for the variance of a type-level quantity}\label{App_stoch_var_eqns}
\input{ind_files/99_Appendix_C}

>>>>

%The abstract goes here. IISER guidelines say it should be <= 250 words.
\thesisabstract{
Population biology is built on a strong mathematical foundation developed during the Modern Synthesis through fields such as theoretical population genetics, evolutionary game theory, and quantitative genetics. Historically, these formalisms have often worked with infinite populations, ignoring the effects of demographic stochasticity. Finite population models in population genetics usually assume a fixed population size and are of limited applicability in the real world, where population sizes routinely fluctuate. In this thesis, I use ideas from statistical physics to analytically describe evolving populations from biological first principles. Starting from a density-dependent ‘birth-death process’ describing an arbitrary closed population of individuals with discrete traits, I derive a set of stochastic differential equations (SDEs) for how trait frequencies change over time. Along with recovering the effects of the standard evolutionary forces of selection, mutation, and drift, these SDEs also reveal a new directional evolutionary force, ‘noise-induced selection’, that is particular to finite populations and has been largely overlooked in standard formulations of evolution. Noise-induced selection can reverse the direction of evolution predicted by infinite-population frameworks, with implications for simulation studies and real world populations. Well-known results such as the replicator-mutator equation and Fisher’s fundamental theorem are recovered in the infinite population limit. Finally, I extend these ideas to one-dimensional quantitative traits through a ‘stochastic field theory’ that yields frameworks such as Kimura's continuum-of-alleles and gradient dynamics in the infinite population limit. My work thus generalizes the formal structures of population biology to finite fluctuating populations and predicts a new evolutionary force unique to such populations.
}

@@

\part{Theory}
\chapter{Population dynamics from stochastic first principles}\label{chap_BD}
\input{ind_files/03_BD_processes}
\chapter{Stochastic field equations for the evolution of quantitative traits}\label{chap_infD_processes}
\input{ind_files/04_field_eqns}

\part{Discussion \& Outlook}
\chapter{A unified view of population dynamics}\label{chap_unification}
\input{ind_files/05_unification_discussion}
%\chapter{Towards a stochastic evolutionary theory}\label{chap_stochasticity}
%\input{ind_files/07_stochastic_discussion}
% \chapter{}
% \input{ind_files/chapter_08}
% \chapter{} <-- uncomment and add as required
% \input{ind_files/chapter_09}

@@

%make sure the title is 'Appendix' and not 'Chapter'
\renewcommand{\chaptername}{Appendix} 
\chapter{Deriving the Fokker-Planck equations for It\^{o} SDEs}\label{App_SDE_FPE}
\input{ind_files/99_Appendix_A}
\chapter{Deriving stochastic trait frequency dynamics using It\^{o}'s formula}\label{App_density_to_freq}
\input{ind_files/99_Appendix_B}
\chapter{A Price-like equation for the variance of a type-level quantity}\label{App_stoch_var_eqns}
\input{ind_files/99_Appendix_C}
\chapter{Some Examples}\label{App_examples}
\input{ind_files/99_Appendix_D_examples}
